---
title: "Mo'orea Temperature Time Series"
format: html
---

## Sampling Timepoints

The temperature data was collected across four key sampling timepoints during the MOSAiC project:

| Timepoint | Date | Season |
|-----------|------|---------|
| TP1 | January 2020 | Austral Summer |
| TP2 | March 2020 | Austral Summer/Autumn |
| TP3 | September 2020 | Austral Winter/Spring |
| TP4 | November 2020 | Austral Spring |

## Interactive Temperature Explorer

This page provides interactive visualization of in-situ temperature logger data from three Mo'orea reef sites.

```{r}
#| echo: false
#| message: false
#| warning: false
#| cache: true
pkgs <- c("vroom","dplyr","lubridate","plotly","here","zoo")
new <- setdiff(pkgs, rownames(installed.packages()))
if (length(new)) install.packages(new)

library(vroom)
library(dplyr)
library(lubridate)
library(plotly)
library(here)
library(zoo)

# Read data (UTC timestamps)
raw <- vroom::vroom(here("data","temperature_timeseries.csv"),
                    col_types = "Tcd", progress = FALSE) |>
  mutate(date.time = ymd_hms(date.time, tz = "UTC"))

# Basic QC: drop impossible temps (<10 or >40 C) if any
raw <- raw |> filter(between(temp.C, 10, 40))

# Precompute aggregations
hourly <- raw |>
  mutate(hour = floor_date(date.time, "1 hour")) |>
  group_by(site, hour) |>
  summarise(temp.C = mean(temp.C), .groups = "drop")

daily <- raw |>
  mutate(day = as_date(date.time)) |>
  group_by(site, day) |>
  summarise(minT = min(temp.C), meanT = mean(temp.C), maxT = max(temp.C), .groups = "drop")

# Rolling (24h) smooth on hourly means
hourly <- hourly |>
  arrange(site, hour) |>
  group_by(site) |>
  mutate(temp_roll24 = zoo::rollapply(temp.C, 24, mean, align = "right", fill = NA)) |>
  ungroup()

# Define sampling timepoints
timepoints <- data.frame(
  timepoint = c("TP1", "TP2", "TP3", "TP4"),
  date = c("2020-01-15", "2020-03-15", "2020-09-15", "2020-11-15"),
  season = c("Austral Summer", "Austral Summer/Autumn", "Austral Winter/Spring", "Austral Spring")
) |>
  mutate(date = ymd(date))

pal <- c(Manava = "#1b9e77", Hilton = "#d95f02", Mahana = "#7570b3")

# Hourly line + optional rolling mean
p_hourly <- plot_ly(type='scatter', mode='lines') |>
  add_trace(data = hourly, x = ~hour, y = ~temp.C, color = ~site, colors = pal,
            hovertemplate = "%{x|%Y-%m-%d %H:%M} UTC<br>%{y:.2f}°C<extra>%{legendgroup}</extra>",
            name = ~site, legendgroup = ~site) |>
  add_trace(data = hourly, x = ~hour, y = ~temp_roll24, color = ~site, colors = pal,
            hovertemplate = "24h mean<br>%{x|%Y-%m-%d %H:%M} UTC<br>%{y:.2f}°C<extra>%{legendgroup}</extra>",
            name = ~paste(site, "(24h mean)"), legendgroup = ~site, line = list(width = 2, dash = 'dash')) |>
  # Add timepoint reference lines
  add_segments(x = ~date, xend = ~date, y = 10, yend = 40, 
               data = timepoints, line = list(color = "red", width = 2, dash = "dot"),
               name = "Sampling Timepoints", showlegend = TRUE) |>
  # Add timepoint labels
  add_annotations(x = ~date, y = 41, text = ~timepoint, 
                  data = timepoints, showarrow = FALSE, 
                  font = list(color = "red", size = 12)) |>
  layout(title = "Hourly Mean Temperature (with 24h Rolling Mean)",
         xaxis = list(title = "Time (UTC)", rangeslider = list(visible = TRUE),
                      rangeselector = list(buttons = list(
                        list(count = 7, label = '7d', step = 'day', stepmode = 'backward'),
                        list(count = 1, label = '1m', step = 'month', stepmode = 'backward'),
                        list(step = 'all')))),
         yaxis = list(title = "Temperature (°C)", 
                     range = c(20, 32),  # Expanded range to show fluctuations better
                     fixedrange = FALSE))  # Allow user to zoom/pan y-axis

# Daily ribbons
p_daily <- plot_ly(type='scatter', mode='lines') |>
  add_ribbons(data = daily, x = ~day, ymin = ~minT, ymax = ~maxT, split = ~site, color = ~site, colors = pal,
              opacity = 0.25, name = ~paste(site, 'Daily Range'), hovertemplate = "%{x|%Y-%m-%d}<br>Min %{y[0]:.2f}°C<br>Max %{y[1]:.2f}°C<extra>%{legendgroup}</extra>") |>
  add_trace(data = daily, x = ~day, y = ~meanT, split = ~site, color = ~site, colors = pal,
            name = ~paste(site, 'Daily Mean'), hovertemplate = "%{x|%Y-%m-%d}<br>Mean %{y:.2f}°C<extra>%{legendgroup}</extra>") |>
  # Add timepoint reference lines
  add_segments(x = ~date, xend = ~date, y = 10, yend = 40, 
               data = timepoints, line = list(color = "red", width = 2, dash = "dot"),
               name = "Sampling Timepoints", showlegend = TRUE) |>
  # Add timepoint labels
  add_annotations(x = ~date, y = 41, text = ~timepoint, 
                  data = timepoints, showarrow = FALSE, 
                  font = list(color = "red", size = 12)) |>
  layout(title = "Daily Min/Mean/Max", 
         yaxis = list(title = "Temperature (°C)", 
                     range = c(20, 32),  # Expanded range to show fluctuations better
                     fixedrange = FALSE))  # Allow user to zoom/pan y-axis

subplot(p_hourly, p_daily, nrows = 2, shareX = FALSE, titleY = TRUE)
```

### Data Download

```{r}
#| echo: false
# Provide a compact table of latest 24h (hourly means) for quick inspection
#latest_window_start <- max(hourly$hour) - hours(24)
#latest <- hourly |> filter(hour >= latest_window_start) |> select(site, hour, temp.C, temp_roll24)

#knitr::kable(latest, caption = "Latest 24 Hours (Hourly Means)")
```

```{r}
#| echo: false
#| eval: false
# Optional example to write processed summaries (disabled by default)
# write.csv(hourly, 'data/temperature_hourly.csv', row.names = FALSE)
# write.csv(daily, 'data/temperature_daily.csv', row.names = FALSE)
```

### Notes
- Raw points aggregated to hourly means for performance.
- 24h rolling mean computed per site to smooth diel variability.
- Timestamps are UTC.

- **Y-axis Features**: 
  - Default range set to 20-32°C for optimal viewing of temperature fluctuations
  - Users can manually zoom and pan the y-axis by dragging or using the mouse wheel
  - Double-click to reset to default view
- **Sampling Timepoints**: Red dotted lines mark TP1-TP4 sampling events for correlation with temperature patterns


### Related Environmental Data

Long term Moorea monitoring data of other environmental characteristics on Moorea's north shore:

Moorea Coral Reef LTER, L. Washburn, A. Brooks, and H. Krumbholz. 2025. MCR LTER: Coral Reef: Gump Station Meteorological Data, ongoing since 2006 ver 53. Environmental Data Initiative. https://doi.org/10.6073/pasta/eccd196e1bf25c97fedb8bd48bfbb4a0

